Activity monitoring assignment - Anastasye
==============================

# Part 1: 
## What is mean total number of steps taken per day?

Loading the data:
```{r}
library(rmarkdown)
activeData <- read.csv("activity.csv", header = TRUE)
```

Total number of steps taken per day:
```{r, echo = TRUE}
stepDay <- aggregate(.~date, data=activeData, sum, na.rm=TRUE)
stepDay
```

Here is a histogram of the data:
```{r, echo = TRUE}
stepTotal <- hist(stepDay$steps, main = "Total number of steps per day", xlab = "Steps per day")
```

The following are the mean and the median of the total number of steps taken per day:

```{r, echo = TRUE}
stepDayMean <- mean(stepDay$steps)
stepDayMean
stepDayMedian <- median(stepDay$steps)
stepDayMedian
```

# Part 2:
## What was the average daily activity pattern?

The average number of steps taken at each 5 minute intervals:
``` {r, echo = TRUE}
stepInterval <- aggregate(steps ~ interval, data = activeData, mean)
```

Here is the time series plot of the steps per 5 minute interval data:
```{r , echo = TRUE}
plot(stepInterval$interval, stepInterval$steps, type = "l", col = 1, main = "Average steps per interval", 
     xlab = "Time intervals", ylab = "Average steps")
```

The following is the interval at which the maximum number of steps were taken:
```{r, echo = TRUE}
MaxInterval <- which.max(stepInterval$steps)
MaxInterval
```

Interval 104 had the highest number of steps taken compare to the rest of the data.

# Part 3:
## Data analysis with missing values included:

Total number of missing values in the dataset:
```{r, echo = TRUE}
sum(is.na(activeData))
```
There are 2304 missing values in the entire data set.

In order to fill in the issing values in the data set, a for loop can be implemented:
```{r, echo = TRUE}
for(i in 1:nrow(activeData)){
    if(is.na(activeData$steps[i])){
        missValue <- stepInterval$steps[which(stepInterval$interval == activeData$interval[i])]
        activeData$steps[i] <- missValue
    }
}
```

Using the loop, a new data set with the missing values was created:
```{r, echo = TRUE}
stepDayNA <- aggregate(steps ~ date, activeData, sum)
```

Here is the histogram of the new data set with the missing values:
```{r, echo = TRUE}
hist(stepDayNA$steps, main = "Total number of steps with NAs", xlab = "Steps per day")
```

Mean and median of the total number of steps taken each day are as followed:
```{r, echo = TRUE}
stepDayNAMean <- mean(stepDayNA$steps)
stepDayNAMean
stepDayNAMedian <- median(stepDayNA$steps)
stepDayNAMedian
```
It appears that the missing values did not affect these calculations. 

# Part 4:
## Were there differences in the activity patterns during weekdays and weekends?

First, a new variable is created to differentiate between weekdays and weekends within the data set:
```{r, echo = TRUE}
ActivePattern <- function(date_val) {
    weekday <- weekdays(as.Date(date_val, '%Y-%m-%d'))
    if  (!(weekday == 'Saturday' || weekday == 'Sunday')) {
        x <- 'Weekday'
    } else {
        x <- 'Weekend'
    }
    x
}
```

Next, this vector is converted into a factor variable:
```{r, echo = TRUE}
activeData$day_type <- as.factor(sapply(stepDayNA$date, ActivePattern))
```

Average number of steps per interval at each corresponding weekday is aggregated together:
```{r, echo = TRUE}
StepsDayNA <- aggregate(steps ~ interval+day_type, activeData, mean)
```

A plot of created to show the comparison in activity patterns between weekday and weekends:
```{r, echo = TRUE}
library(ggplot2)
plot <- ggplot(StepsDayNA, aes(interval, steps)) +
    geom_line(stat = "identity", aes(colour = day_type)) +
    theme_gray() +
    facet_grid(day_type ~. , scales="fixed", space="fixed") +
    labs(x="Interval", y="Number of Steps", title = "Number of steps by day type")
    
print(plot)
```
It appears that weekends had a slightly higher activity patters, meaning more steps taken per day
compared to weekdays.